---
title: "Fire Incident Analysis"
subtitle: "Exploring Toronto Fire Incident Data"
author: 
  - Bruce Yu
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "In this paper, we analyze fire incident data in Toronto to identify patterns and relationships between response times, estimated losses, and various incident characteristics. Using visualization techniques, we provide insights into key metrics such as incident type, response time, and resource allocation."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)

cleaned_data <- read_csv("../data/cleaned_data/fire_incidents_cleaned.csv")
```


# Introduction

Fire safety is a critical concern in urban environments, where the density of buildings and population heightens the risks of fire-related incidents. In major cities like Toronto, understanding the dynamics of fire incidents, response times, and the associated economic losses is essential for policymakers and emergency responders to optimize their strategies. Although fire department interventions and fire alarm systems are in place to mitigate risks, there is a need for data-driven analyses that shed light on how these factors interact and impact overall fire safety outcomes.

This paper focuses on the analysis of fire incidents in Toronto, drawing from a dataset provided by the City of Toronto’s Open Data initiative. The dataset contains detailed records of fire incidents, response times, estimated financial losses, and fire alarm system statuses. Previous studies have largely focused on the importance of timely responses and fire alarm systems in minimizing fire damage. However, there remains a gap in quantifying the direct relationship between response times, fire alarm presence, and the financial impact of fire incidents. Additionally, little attention has been given to spatial disparities in fire incidents across different areas of the city, particularly how station areas might experience varying frequencies of incidents.

To address these gaps, we conducted a comprehensive analysis of fire incidents in Toronto, examining how response times and the presence of fire alarm systems influence the financial losses incurred from fires. We also explored the distribution of fire incidents across different station areas, seeking to identify regions with higher fire risks. Using statistical methods, we visualized the distributions and relationships between key variables such as response times, estimated financial losses, and fire alarm system efficacy.

Our findings suggest that, while shorter response times generally correlate with lower estimated losses, there are notable exceptions, especially in incidents where fire alarms were either absent or ineffective. Certain station areas also exhibited higher frequencies of fire incidents, indicating possible regional risk factors that warrant further investigation. These insights are critical for enhancing fire safety protocols, particularly in resource allocation and public awareness campaigns targeting fire alarm installations.

The remainder of the paper is structured as follows. Section 2 provides a detailed overview of the dataset, including its source, key variables, and any preprocessing steps. Section 3 presents the results of our analyses, including visualizations of response times, estimated losses, and station area distributions. Section 4 discusses the implications of these findings, particularly in relation to policy and future research. Finally, Section 5 offers concluding remarks and suggests potential next steps for improving fire safety measures based on our findings.

The introduction is self-contained and tellsa reader everything they need to knowincluding: 1) broader context to motivate;2) some detail about what the paper isabout; 3) a clear gap that needs to befilled; 4) what was done; 5) what wasfound; 6) why it is important; 7) thestructure of the paper. A reader should beable to read only the introduction andknow what was done, why, and what wasfound. Likely 3 or 4 paragraphs, or 10 percent of total.

The remainder of this paper is structured as follows. @sec-data provides a detailed description of the data used. In @sec-results, we visualize the results, and @sec-discussion discusses the findings.


# Data {#sec-data}

The dataset used in this analysis contains over 32,000 records from Toronto fire incidents. It includes various features such as response times, estimated losses, number of responding units, presence of fire alarms, and more. Each fire incident is categorized by incident type, and there are records for both civilian and firefighter casualties.

The following sections explore different aspects of this dataset using various visualization techniques.

### Distribution of Response Time

The distribution of response times for fire incidents is shown in @fig-response-time. The majority of incidents have response times between 4 and 6 minutes, with a few incidents having much longer response times.


```{r}
#| label: fig-response-time
#| fig-cap: "Distribution of Response Time"
ggplot(cleaned_data, aes(x = Response_Time)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Response Time", x = "Response Time (minutes)", y = "Count") +
  theme_minimal()
```

## Distribution of Estimated Loss
The distribution of estimated loss for fire incidents is highly skewed, as shown in @fig-estimated-loss. Most incidents report estimated losses below $5,000, though a small number of incidents have significantly higher losses, up to $50,000 or more.

```{r}
#| label: fig-estimated-loss
#| fig-cap: "Distribution of Estimated Loss"
ggplot(cleaned_data, aes(x = Estimated_Loss)) +
  geom_histogram(binwidth = 5000, fill = "lightgreen", color = "black") +
  labs(title = "Distribution of Estimated Loss", x = "Estimated Loss ($)", y = "Count") +
  theme_minimal()
```

## Response Time by Incident Type
To examine how response times vary across different types of incidents, we plotted a boxplot of response times grouped by incident type in @fig-response-incident-type. It shows that the median response time is relatively similar across most incident types, with a few outliers where response times are much longer.

```{r}
#| label: fig-response-incident-type
#| fig-cap: "Response Time by Incident Type"
# Simplify the Incident_Type column
cleaned_data <- cleaned_data %>%
  mutate(Incident_Type_Simplified = case_when(
    grepl("01", Incident_Type) ~ "Fire",
    grepl("02", Incident_Type) ~ "Explosion",
    grepl("03", Incident_Type) ~ "No Loss Outdoor Fire",
    TRUE ~ "Other"
  ))

# Plot using the simplified names
ggplot(cleaned_data, aes(x = Incident_Type_Simplified, y = Response_Time)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(title = "Response Time by Incident Type", x = "Incident Type", y = "Response Time (minutes)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


```

## Response Time vs Estimated Loss
We also explored the relationship between response time and estimated loss. The scatter plot in @fig-response-estimated-loss suggests that there is a weak positive correlation between response time and estimated loss. However, the trend is not strong, and there are a large number of incidents with zero or low losses regardless of response time.

```{r}
#| label: fig-response-estimated-loss
#| fig-cap: "Response Time vs Estimated Loss"
ggplot(cleaned_data, aes(x = Response_Time, y = Estimated_Loss)) +
  geom_point(alpha = 0.5, color = "darkblue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Response Time vs Estimated Loss", x = "Response Time (minutes)", y = "Estimated Loss ($)") +
  theme_minimal()

```


## Number of Fire Incidents by Station Area
The number of fire incidents by station area is visualized in @fig-fire-incidents-station-area. The bar plot shows the number of fire incidents by station area, with the station areas segmented into four distinct groups on the x-axis (around 100, 200, 300, and 400). Each group represents different geographic regions or station numbers in the dataset.

Station areas in the 300 range show the highest density of fire incidents, with several exceeding 750 incidents. This might indicate that these regions are more populated or prone to higher fire risks.
Other station areas (around 100, 200, and 400) also see fire incidents, but their distribution is less dense compared to the 300 series.
The clustering of incidents around certain station areas could be driven by multiple factors, such as population density, industrial areas, or geographical layouts prone to fire hazards. Further investigation is needed to determine whether these station areas correspond to specific urban or suburban regions.

```{r}
#| label: fig-fire-incidents-station-area
#| fig-cap: "Number of Fire Incidents by Station Area"
# Remove rows with non-finite values in the Incident_Station_Area column
cleaned_data_filtered <- cleaned_data %>%
  filter(!is.na(Incident_Station_Area) & is.finite(Incident_Station_Area))

# Re-plot the data after filtering
ggplot(cleaned_data_filtered, aes(x = Incident_Station_Area)) +
  geom_bar(fill = "orange") +
  labs(title = "Number of Fire Incidents by Station Area", 
       x = "Station Area", 
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```
## Responding Units vs Estimated Loss
The relationship between the number of responding units and estimated loss is shown in @fig-responding-units-loss. There appears to be a positive correlation between the number of units and the estimated loss, which may indicate that larger incidents with higher estimated losses require more resources.

```{r}
#| label: fig-responding-units-loss
#| fig-cap: "Responding Units vs Estimated Loss"
ggplot(cleaned_data, aes(x = Responding_Units, y = Estimated_Loss)) +
  geom_point(alpha = 0.5, color = "purple", position = position_jitter(width = 0.2, height = 500)) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Responding Units vs Estimated Loss", x = "Number of Responding Units", y = "Estimated Loss ($)") +
  theme_minimal() +
  xlim(0, 50)

```


# References



